Matching Bullets

### 2015-12-19 imaging-paper.Rnw
update on random forest

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we ues cross sections 25$\mu m$ apart and require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface.}
%
\begin{figure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\linewidth]{images/br6-2-1.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br9 Bullet 2-4}
\includegraphics[width=\linewidth]{images/br9-2-4.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet B-2}
\includegraphics[width=\linewidth]{images/b-2.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet Q-4}
\includegraphics[width=\linewidth]{images/q-4.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all unknown bullets against all known bullets for matches. 
The algorithm is using land-to-land matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the number of maximum consecutive matching striae between land C-3 and all 120 known lands. Two lands show a high CMS. These correspond to the known matches, shown in Figures~\ref{fig:hamby-perfect}b and c. 
 Unfortunately, not all results are as clear cut. 
It is not reasonable to assume that we can match all lands, but we can try to maximize the number of matches to get an overview of what we can reasonably expect from an automated match.

@@

bstats$pred <- predict(rp1, newdata=bstats)
@
\caption{\label{fig:tree}Decision tree of matching bullets based on recursive partitioning. The rectangular nodes are the leaves, giving a short summary consisting of the number of observations in the leaf (bottom left), the corresponding percentage of the total (bottom right). The number at the top shows the fraction of these observations that are a match. A 1 or a 0 therefore indicate a homogeneous (or perfect) node. }
\end{figure}

@@

errors <- data.frame(rtrees$err.rate)
errors$id <- 1:nrow(errors)
bstats$forest <- predict(rtrees, type="prob", newdata=bstats)[,2]
@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.7\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0044. This is a weighted average between zero false positives and an error rate of false negatives of 0.26. Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.7\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))

bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
qplot(data=bm, value, reorder(bullet, value/n), size=I(2.5), 
      shape=prediction, colour=prediction, alpha=I(0.6)) + 
  theme_bw()  + ylab("") + 
#  scale_shape_manual(values=c(1,2)) +
  scale_x_continuous("Number of correctly predicted land-to-land matches", 
                     breaks = 3*0:4, limits=c(0,12),
                     minor_breaks = 1:11) +
  theme(legend.position = "bottom")
@
\caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. For each bullet at least four lands are correctly matched. Unknown bullet B is the hardest for the algorithm to match.}
\end{figure}
%
However, this does not carry a lot of weight because of the few bullet comparisons in question: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger test case. 
Another thing to note about the  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on 0.5 is not necessary. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we can therefore introduce an interval of `inconclusive' results in the middle of the spectrum. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree ($x$-axis) and the random forest ($y$-axis). Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

\begin{figure}[hbtp]
<<tree:forest, echo=FALSE, fig.width=6, fig.height=6.25, out.width='.7\\linewidth'>>=
qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
      shape=match) + theme_bw() +
  theme(legend.position="bottom") +
  geom_vline(xintercept=0.5, colour="grey50") +
  geom_hline(yintercept=0.5, colour="grey50") + 
  xlab("Predictions from decision tree") +
  ylab("Predictions from random forest")   +
  scale_colour_brewer(palette="Paired")
@
\caption{\label{fig:tree-forest}Scatterplot of prediction results from the tree (horizontal) and the forest (vertical). Each point corresponds to a land-to-land comparison. Known matches are indicated in blue. Matches predicted by the decision tree are on the right half of the plot, matches predicted by the random forest are in the upper half. }
\end{figure}

>>>>

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we ues cross sections 25$\mu m$ apart and require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface.}
%
\begin{figure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\linewidth]{images/br6-2-1.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Br9 Bullet 2-4}
\includegraphics[width=\linewidth]{images/br9-2-4.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet B-2}
\includegraphics[width=\linewidth]{images/b-2.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}
\caption{Unkn Bullet Q-4}
\includegraphics[width=\linewidth]{images/q-4.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the number of maximum consecutive matching striae between land C-3 and all 120 known lands. Two lands show a high CMS. These correspond to the known matches, shown in Figures~\ref{fig:hamby-perfect}b and c. 
 Unfortunately, not all results are as clear cut. 
It is not reasonable to assume that we can match all lands, but we can try to maximize the number of matches to get an overview of what we can reasonably expect from an automated match.

@@

bstats$pred <- predict(rp1, newdata=bstats)
bstats$pred[bstats$flagged] <- NA
@
\caption{\label{fig:tree}Decision tree of matching bullets based on recursive partitioning. The rectangular nodes are the leaves, giving a short summary consisting of the number of observations in the leaf (bottom left), the corresponding percentage of the total (bottom right). The number at the top shows the fraction of these observations that are a match. A 1 or a 0 therefore indicate a homogeneous (or perfect) node. }
\end{figure}

@@

errors <- data.frame(rtrees$err.rate)
errors$id <- 1:nrow(errors)
bstats$forest <- predict(rtrees, type="prob", newdata=bstats)[,2]
bstats$forest[bstats$flagged] <- NA
@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.7\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0040. This is a weighted average between a false positive error rate of 0.000098 and an error rate of false negatives of 0.22. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known matches and non-matches (see Figures~\ref{fig:treeresults} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.7\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))

bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
      shape=prediction, colour=prediction, alpha=I(0.6)) + 
  theme_bw()  + ylab("") + 
#  scale_shape_manual(values=c(1,2)) +
  scale_x_continuous("Percent of correctly predicted land-to-land matches", 
                     breaks = c(0,25,50,75,100), limits=c(0,100)) +
  theme(legend.position = "bottom")
@
\caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
\end{figure}
%
However, this does not carry a lot of weight because of the few bullet comparisons in question: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger test case. 
Another thing to note about the  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on 0.5 is not necessary. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we can therefore introduce an interval of `inconclusive' results in the middle of the spectrum. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree ($x$-axis) and the random forest ($y$-axis). Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

\begin{figure}[hbtp]
<<tree:forest, echo=FALSE, fig.width=6, fig.height=6.25, out.width='.7\\linewidth'>>=
qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
      shape=match) + theme_bw() +
  theme(legend.position="bottom") +
  geom_vline(xintercept=0.5, colour="grey50") +
  geom_hline(yintercept=0.5, colour="grey50") + 
  xlab("Predictions from decision tree") +
  ylab("Predictions from random forest")   +
  scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
  scale_shape_discrete(labels=c("KNM", "KM"))
@
\caption{\label{fig:tree-forest}Scatterplot of prediction results from the tree (horizontal) and the forest (vertical). Each point corresponds to a land-to-land comparison. Known matches are indicated in blue. Matches predicted by the decision tree are on the right half of the plot, matches predicted by the random forest are in the upper half. }
\end{figure}
